---
title: "Network Analysis of the Enslaved Community at Mount Vernon, 1786-1799"
author: "Sara Collini"
date: "December 10, 2014"
output: html_document
---

#Introduction
This project explores the kinship relations of the Mount Vernon enslaved community from 1786-1799. Currently, there is great interest in studying the enslaved community on George Washington's estate, such as archaeology of the slave cemetery, the physical spaces in which the slaves lived, their labor, free time, and Washington’s role as a slave owner. The Fred W. Smith National Library for the Study of George Washington staff is currently building a slavery database, in which every mention of every slave that lived, worked, or had some part at Mount Vernon is entered. The documents include farm reports, wills, letters, diaries, censuses, etc. In order to track all of the unique people, they generated a spreadsheet compiling every census taken of the slaves. 

The census is relatively complete, with farm location, estimated age, skill, spouses, and mothers. This project draws from that data in order to explore the kinship relations from the two most complete censuses: 1786 and 1799. We can parse out parent/child, sibling, and spousal relationships. This project examines the network of known kinship relationships on the estate, the overall structure of the enslaved community, and how it changes in those 13 years. Hopefully, a better and more complete picture of the Mount Vernon enslaved community will arise. 

HISTORIOGRAPHY

These slave kinship relations are difficult to study, because the source base is so minimal. If plantation censuses do exist, they are usually fragmentary. Because Washington kept a relatively complete record of his slaves' information, we can learn quite a bit from studying the kinship network and community structure of the Mount Vernon slaves. 

A note on definitions/ terms: Slaves could not legally marry. A “husband” or “wife” listed on the census does not necessarily mean the same thing as a legal marriage, nor does it imply behaviors associated with legal marriage. A structure of two parents or a single parent with children does not necessarily constitute a traditional family, as perceived by a slave. Additionally, polygyny and polyandry were common. We cannot assume blood relations or a slave’s perspective when no direct sources exist. Hoever, in order to analyze the community, I have decided to call the ties “kinship or familial relationships” that make up “family structures”.

First, we load the necessary libraries.
```{r}
library(igraph)
library(dplyr)
library(tidyr)
library(stringr)
library(historydata)
library(ggplot2)
```

#Data Manipulation

##Read in the data
The data for this project exists in three csv files. The first file contains every slave with at least one kinship relation on Mount Vernon, from 1786-1799
.
```{r}
slaverel <- read.csv("~//Desktop/Clio-3/finalproject/slaveRelationships.csv", stringsAsFactors = FALSE)
head(slaverel)
```

The second contains every slave and the respective static information, such as gender and owner, for 1786 and 1799.
```{r}
total_slave <- read.csv("~//Desktop/Clio-3/finalproject/clean.csv", stringsAsFactors = FALSE) %>%
  select(id, Gender, Birth.Year, Owner)
head(total_slave)
```

This file contains both the 1786 and 1799 censuses with the relevant slaves. This means the slaves present in both censuses are repeated but might have a different skill or farm. 
```{r}
slave_census <- read.csv("~//Desktop/Clio-3/finalproject/census.csv", stringsAsFactors = FALSE) %>%
  select(id, Gender, Birth.Year, Skill, Farm, Census, Owner)
head(slave_census)
```

##Data Joining 
Next, we must separate out the two most important and complete censuses: 1786 and 1799. We select only the variables that change between the two censuses. 
```{r}
slave_1786 <- slave_census %>%
  filter(Census == "1786") %>%
  select(id, Census, Skill, Farm)

slave_1799 <- slave_census %>%
  filter(Census == "1799") %>%
  select(id, Census, Skill, Farm)
```


Join the two censuses into one dataframe by the "id" variable in order to access that information on the network graph. 
```{r}
total_slave <- total_slave %>%
  left_join(slave_1786, by = "id") %>%
  left_join(slave_1799, by = "id")
head(total_slave)
```

#Create a Graph Object
We create a graph object of the slaverel dataframe in order to graph the kinship relations between the slaves.
```{r}
slave_graph <- graph.data.frame(slaverel, directed = "FALSE")
```

In order to add those slaves without kinship relations, we must add them as extra vertices to the graph object. 
```{r}
slave_graph <- slave_graph + vertex("Adam B", "Adam C", "Anthony A", "Austin A", "Bath A", "Bristol A", "Brunswick A", "Caesar A", "Cloe A", "Cupid A", "Daniel A", "Daphne D", "Doll A", "Doll E", "Essex A", "Frank B", "Giles A", "Hannah E", "Hercules A", "Isaac B", "Jack B", "Jack G", "Jack H", "Jack I", "James D", "Jenny B", "Jenny E", "Joe E", "Juba A", "Julius A", "London A", "Marcus A", "Matt A", "Milly C", "Moll A", "Molly A", "Morris B", "Moses C", "Murria A", "Nancy I", "Paris A", "Paul A", "Peter C", "Robin A", "Robin B", "Schomberg A", "Spencer A", "Sue A", "Tom D", "William Lee", "Will D", "Will F", "Will H")
```

Now, we create another dataframe that lists all the vertex attributes for each slave in order to visualize them on the network graph.
```{r}
slave_vertex <- data_frame(name = V(slave_graph)$name) %>%
  left_join(total_slave, by = c("name" = "id"))
head(slave_vertex)
```

#Plotting the First Network Graph
```{r}
igraph.options(vertex.size=3, 
               edge.arrow.size=.5)
set.seed(5)
plot(slave_graph, vertex.label = NA)
title("Slave Families on Mount Vernon, 1786 - 1799")
```
Immediately, the overall kinship structure of the enslaved community is apparent from this network graph. Four family groups comprise one large familial relation and several smaller groups also exist. Many slaves had no kinship relations (no known kinship relations) and some slaves were only connected to one or two other people. In order to analyze this information further, we must be able to visualize the vertex attributes and the different kinship relationships.

##Gender
In order to visualize the gender of the slaves, we can change the shape of each vertex accordingly.
```{r}
V(slave_graph)$shape <- ifelse(slave_vertex$Gender == "Female", "circle", 
                               ifelse(slave_vertex$Gender == "Male", "square", 
                                      "rectangle"))
set.seed(5)
plot(slave_graph, vertex.label = NA)
title("Visualizing Gender")
```
Females are circles, while males are squares. 

##Relationships
In order to visualize the different kinships relations, we can colorize the edges according to type of relation with a function.
```{r}
lookup_edge <- function(relationship) {
  if(relationship == "Spouse") return("red")
  if(relationship == "Child") return("orange")
  if(relationship == "Sibling") return("blue")
  return("gray")
} 

lookup_edge <- Vectorize(lookup_edge, USE.NAMES = FALSE)

E(slave_graph)$color <- lookup_edge(E(slave_graph)$relationship)
set.seed(5)
plot(slave_graph, 
     vertex.label = NA,      
     vertex.size = 3)
title("Visualizing Kinship Relations")
legend("bottomleft", legend = c("Spouse", "Child", "Sibling"),
       col = c("red", "orange", "blue"), pch = 19,   
               title = "Kinship Relation")
```
The network graph is easier to decipher. We can see that the one large family is connected by a slave marriage. Many of the two person families are slave marriages, while others are mother and child. With this graph, we can count the number of each relationship and know where they exist within each family. 


#Evolution of Kinship Networks in Space
Since both slave censuses taken by Washington list the farms that each slave lived and worked on, we can use that variable to locate the slave families within the estate. We use a function similar to the edge coloring function:
```{r}
lookup_color <- function(type) {
  if(is.na(type)) return("gray")
  if(type == "Mansion House") return("purple")
  if(type == "Muddy Hole") return("blue")
  if(type == "Dogue Run") return("green")
  if(type == "Union Farm") return("orange")
  if(type == "River Farm") return("yellow")
  if(type == "Mr. Lears") return("brown")
  if(type == "Mrs. Washington's") return("brown")
  if(type == "Cedar Grove") return("brown")
  return("gray")
} 

lookup_color <- Vectorize(lookup_color, USE.NAMES = FALSE)
```
Five Farms made up the Mount Vernon estate: Mansion House, Muddy Hole, Dogue Run, Union Farm, and River Farm. Most slaves married and had kinship relations within Washington's slave community. However, some slaves married people owned by others. I have colored those non-Mount Vernon slave spouses brown. The gray color represents slaves that were not active on each census, meaning that they were not born yet or had died. 

##Slave Families Across Farms in 1786
```{r}
V(slave_graph)$color <- lookup_color(slave_vertex$Farm.x)
set.seed(5)

plot(slave_graph, 
     vertex.label = NA,      
     vertex.size = 3)
title("Slave Families on Mount Vernon Farms in 1786")
legend("bottomleft", legend = c("Mansion House", "Muddy Hole", "Dogue Run", "Union Farm", "River Farm", "Other Farm", "Not Active"),
       col = c("purple", "blue", "green", "orange", "yellow", "brown", "gray"), pch = 19,   
               title = "Farm")
```
This graph tells us a lot more information. We can place the slave families in space on the estate and organize the various family structures. 

###1786 Family Structures 
From this graph, four types of family structures arise: singles, one parent, two parent, and couples with no known children. We can plot the numbers of each family structure for the entire estate, as well as the individual farms. We will use ggplot2 bar graphs to plot this information drawn from the network graph into a csv.

```{r}
structures <- read.csv("~//Desktop/Clio-3/finalproject/FamilyStructures.csv", stringsAsFactors = FALSE)
head(structures)
```

We can use a simple bar plot to show the percentages of each family structure:
```{r}
structures_1786 <- structures %>%
  filter(Census == "1786")

ggplot(data = structures_1786, aes(x = Family.structure, y = Percent)) + facet_wrap(~Farm) + geom_bar(stat = "identity", fill = "darkblue") + theme(axis.text.x=element_text(angle = 90, hjust = 0)) + ggtitle("1786 Family Structures")
```
From this plot, we can see that almost every farm and the entire estate had a majority of single person structures. 

##Slave Families Across Farms in 1799
```{r}
V(slave_graph)$color <- lookup_color(slave_vertex$Farm.y)
set.seed(5)

plot(slave_graph, 
     vertex.label = NA,      
     vertex.size = 3)
title("Slave Families on Mount Vernon Farms in 1799")
legend("bottomleft", legend = c("Mansion House", "Muddy Hole", "Dogue Run", "Union Farm", "River Farm", "Other Farm", "Not Active"),
       col = c("purple", "blue", "green", "orange", "yellow", "brown", "gray"), pch = 19,   
       title = "Farm")
```

###1799 Family Structures
Let's plot the four family structures for 1799 next to the 1786 figures:
```{r}
ggplot(data = structures, aes(x = Family.structure, y = Percent, fill = Census)) + facet_wrap(~Farm) + geom_bar(stat = "identity", position = "dodge") + theme(axis.text.x=element_text(angle = 90, hjust = 0)) + ggtitle("1786 and 1799 Family Structures Compared")
```
Immediately, we see the increase in the percentage of two-parent family strucutres in just 13 years. Single person structures still dominate on Mansion House and Union Farm, but every farm and the overall estate increases in two-parent families. 

#Community Detection
We can use the community detection methods in R in order to focus on each family grouping in 1799.
```{r}
comm <- walktrap.community(slave_graph)
length(comm)
```
The walktrap.community DESCRIBE and calculates 119 slave family groups on Mount Vernon in 1799. This is not completely accurate, as the largest family is counted as 4 groups. A more nuanced analysis of the number and definition of family groupings is needed for this data, but the exploring the individual groups is helpful. 

We can also calculate the size of each of the family groups.
```{r}
sizes(comm)
```

We can inspect each community in the graph and list the names of each family member.
```{r}
memb <- membership(comm)
names(memb[memb ==4])
```

We can also show the kinship network graph colored by each family group.
```{r}
set.seed(5)
plot(slave_graph, vertex.label = NA, vertex.color = memb, vertex.size = 3)
title("Families detected by walktrap.community()")
```

```{r}
set.seed(5)
plot(comm, slave_graph, vertex.label = NA, vertex.size = 2)
title("Families detected by walktrap.community()")
```

##Family Tree
In order to see how each slave is connected within a family, we can show a subgraph of the larger network graph in a different layout. We will use community 2 for this example.
```{r}
induced.subgraph(slave_graph, which(membership(comm) %in% c(4,6,10))) %>% 
  plot()
set.seed(20)
title("Communities 4, 6, and 10 among Mount Vernon")
legend("bottomleft", legend = c("Spouse", "Child", "Sibling"),
       col = c("red", "orange", "blue"), pch = 19,   
       title = "Relationship")
```
This graph is not the most visually appealing, nor does it show a true family tree in chronological and hierarchical order. Suckey Bay and NA9 are married and have three children: Nancy G, Nancy H, and Rose A. Rose A has four childern of her own: Hagar A, Joe D, Simon A, and Tom C. This entire family does reside on River Farm in 1799, except Suckey Bay's spouse, who is unknown. This subgraph does show a two generational family living and working on the same farm in 1799. They are all connected by their familial ties, as well as their working community.  

Let's explore a smaller family
```{r}
induced.subgraph(slave_graph, which(membership(comm) == 32)) %>%
  plot(layout = layout.reingold.tilford, vertex.label = names(memb[memb == 32]))
title("Community 32 among Mount Vernon")
legend("bottomleft", legend = c("Spouse", "Child", "Sibling"),
       col = c("red", "orange", "blue"), pch = 19,   
       title = "Relationship")
```
This graph shows that Daphne B and Simms A are spouses with three children: Maria A, Matilda B, and Polly A. The women live on Union Farm, but Simms A lives on Mansion House farm. This means that Daphne B and Simms A would have traveled across these two farms in order to see each other. Simms A belongs to two different communities: his family and the Mansion House working community. 

#Summary Statistics

##Marriages
```{r, eval=FALSE}
marriages <- slaverel %>%
  filter(relationship == "Spouse")

slave_graph_marriages <- graph.data.frame(marriages, directed = "FALSE")

slave_marriages_vertex <- data_frame(name = V(slave_graph_marriages)$name) %>%
  left_join(total_slave, by = c("name" = "id"))

V(slave_graph_marriages)$shape <- ifelse(slave_marriages_vertex$Gender == "Female", "circle", 
                               ifelse(slave_marriages_vertex$Gender == "Male", "square", 
                                      "rectangle"))

E(slave_graph_marriages)$color <- lookup_edge(E(slave_graph_marriages)$relationship)
V(slave_graph_marriages)$color <- lookup_color(slave_marriages_vertex$Farm.y)
plot(slave_graph_marriages, vertex.label = NA) 
title("Slave Marriages on Mount Vernon, 1799")
```

```{r, eval=FALSE}
mar <- walktrap.community(slave_graph_marriages)
length(mar)
membmar <- membership(mar)
names(membmar[membmar ==1])
induced.subgraph(slave_graph_marriages, which(membership(mar) == 1)) %>% plot()
```

##Children
```{r}
children_1786 <- slave_1786 %>%
  filter(Skill == "Child") %>%
  group_by(Farm) %>%
  summarize(Number = n()) %>%
  mutate(Percent_1786 = Number/89 * 100)

children_1799 <- slave_1799 %>%
  filter(Skill == "Child") %>%
  group_by(Farm) %>%
  summarize(Number = n()) %>%
  mutate(Percent_1799 = Number/117 * 100) %>%
  left_join(children_1786, by = "Farm") %>%
  gather(census, percent, Percent_1786,Percent_1799)

ggplot(data = children_1799, aes(x = Farm, y = percent, fill = census)) + geom_bar(position = position_dodge(), stat = "identity") + theme(axis.text.x=element_text(angle = 90, hjust = 0)) + ggtitle("Ratio of Children on Mount Vernon Farms, 1786-1799")
```

##Gender Ratios 
```{r}
gender_1786_females <- slave_vertex %>%
  filter(Census.x == "1786") %>%
  filter(Skill.x != "Child") %>%
  filter(Gender == "Female") %>%
  group_by(Farm.x) %>%
  summarize(Females = n()) 

gender_1786_males <- slave_vertex %>%
  filter(Census.x == "1786") %>%
  filter(Skill.x != "Child") %>%
  filter(Gender == "Male") %>%
  group_by(Farm.x) %>%
  summarize(Males = n())

gender_1786 <- gender_1786_females %>%
   left_join(gender_1786_males, by = "Farm.x") %>%
  mutate(Total = Females + Males) %>%
  mutate(Percent_of_Male = Males/Total * 100) %>%
  mutate(Percent_of_Female = Females/Total * 100) %>%
  gather(gender, percent, Percent_of_Male:Percent_of_Female)

gender_1786_plot <- ggplot(data = gender_1786, aes(x = Farm.x, y = percent, fill = gender)) + geom_bar(position = position_dodge(), stat = "identity") + theme(axis.text.x=element_text(angle = 90, hjust = 0)) + ggtitle("Gender Ratio on Mount Vernon Farms in 1786")

```

```{r}
gender_1799_females <- slave_vertex %>%
  filter(Census.y == "1799") %>%
  filter(Skill.y != "Child") %>%
  filter(Gender == "Female") %>%
  filter(Farm.y != "NA") %>%
  group_by(Farm.y) %>%
  summarize(Females = n()) 

gender_1799_males <- slave_vertex %>%
  filter(Census.y == "1799") %>%
  filter(Skill.y != "Child") %>%
  filter(Gender == "Male") %>%
  filter(Farm.y != "NA") %>%
  group_by(Farm.y) %>%
  summarize(Males = n())

gender_1799 <- gender_1799_females %>%
   left_join(gender_1799_males, by = "Farm.y") %>%
  mutate(Total = Females + Males) %>%
  mutate(Percent_of_Male = Males/Total * 100) %>%
  mutate(Percent_of_Female = Females/Total * 100) %>%
  gather(gender, percent, Percent_of_Male:Percent_of_Female)

gender_1799_plot <- ggplot(data = gender_1799, aes(x = Farm.y, y = percent, fill = gender)) + geom_bar(position = position_dodge(), stat = "identity") + theme(axis.text.x=element_text(angle = 90, hjust = 0)) + ggtitle("Gender Ratio on Mount Vernon Farms in 1799")
```
